---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


``` {r setoptions, echo = FALSE}
library(knitr)
opts_chunk$set(echo = TRUE)
```

## Downloading the data

The data can be downloaded from [here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip).

This code gets the data downloaded and stored in file `activity.csv` in the working directory.

``` {r gettingData}

if(!file.exists("activity.csv")){
        zipfileURL = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
        download.file(zipfileURL, destfile="activity.zip", method = "libcurl")
        unzip("activity.zip")
}
```


## Loading and preprocessing the data

The data in the `activity.csv` file is load into R. Then we take a fist look at the data, using the head, tail and str functions of R.

``` {r loadingAndFirstLook}

activityData = read.csv("activity.csv")
head(activityData)
tail(activityData)
str(activityData)
summary(activityData)
```
 
 
It can be seen that:

- The `steps` variable has missing values (denoted by `NA`), and recognized as integer type 
- The `date` variable is recognized as a factor so it needs to be transformed into a Date type
- The `interval` variable is recognized as an integer type with no missing values.
 
 Getting the `date` variable converted in date format.
 
``` {r processingData1}
activityData$date = as.Date(activityData$date) 
summary(activityData$date)
str(activityData$date)

``` 
 
The variable `date` is coerced into the Date type and checked for missing values.

## What is mean total number of steps taken per day?

Firts, we take a look at a histogram or the total number of steps taken per day.
Second, we see a summary of the daily total number of steps statistics 

``` {r question 1, results = 'asis'}

stepsPerDay = data.frame(stepsPerDay = with(activityData, tapply(steps, date, function(x) sum(x, na.rm= TRUE))))
library(ggplot2)
g1 <- ggplot(data = stepsPerDay, aes(stepsPerDay))
g1 <- g1 + geom_histogram()+labs(x="Total number of steps in a day", y = "Frequency", title = "Histogram")
g1
library(xtable)
summSteps <- summary(stepsPerDay$stepsPerDay)
xt <- xtable(t(summSteps))
caption(xt) = "Statistics for Daily Total Number of Steps"
print(xt, type = "html",include.rownames = FALSE)
```

The median is `r formatC(summSteps[3], format = "d", big.mark = ",")` daily total steps and the mean is `r formatC(summSteps[4], format = "d", big.mark = ",")` daily total steps. 

## What is the average daily activity pattern?

``` {r question 2, results = 'asis'}


meanStepsInterval = data.frame(mean = with(activityData, 
                                           tapply(steps, interval, 
                                                  function(x) mean(x, na.rm = TRUE))))
plot(rownames(meanStepsInterval),meanStepsInterval$mean,type="l", main="Average Number of Daily Steps per 5-minute Interval", xlab = "5-minute Interval", ylab="" )
```

```{r question2.2}

# Interval with the maximum number of steps

intervalMaxSteps = as.integer(names(which.max(meanStepsInterval$mean)))
maxAverSteps = as.numeric(meanStepsInterval$mean[which.max(meanStepsInterval$mean)])
intervalPrev = intervalMaxSteps - 5
intervalHours1 = as.integer(intervalMaxSteps/60)
intervalMinutes1 = intervalMaxSteps - intervalHours1*60
intervalHours2 = as.integer(intervalPrev/60)
intervalMinutes2 = intervalPrev - intervalHours2*60
```

The maximum average number of steps in a 5-min interval is `r format(maxAverSteps, digits = 4)` and it occurs between  `r paste(intervalHours2,":",intervalMinutes2, sep ="")` and `r paste(intervalHours1,":",intervalMinutes1, sep = "")`.

## Imputing missing values

```{r missing}

goodRows = complete.cases(activityData)
numberRowsMissing = length(activityData$steps) - sum(goodRows)
numberRowsMissingPct = numberRowsMissing / length(activityData$steps) *100
colSums(is.na(activityData))
```


The total number of rows with missing values is `r formatC(numberRowsMissing, format = "d", big.mark = ",")`, which is a
`r paste(format(numberRowsMissingPct, digits = 3), "%", sep = "")` of the total number of rows.  This number coincides with the number of missing values for the steps variable.

```{r missing2}

activityData2 = activityData
meanStepsInterval$interval = as.integer(rownames(meanStepsInterval$mean))
for (i in 1:length(goodRows)) {
        if(!goodRows[i]) {
                activityData2$steps[i] = meanStepsInterval$mean[which(meanStepsInterval$interval == 
                                                                      activityData2$interval[i])]
        }
}
summary(activityData2)
```


These missing values were substituted by the mean number steps for the corresponding 5-minute interval.  A new dataset was created with the missing data fill in this way.

``` {r histogram2, results = 'asis'}

stepsPerDay2 = data.frame(stepsPerDay = with(activityData2, tapply(steps, date, sum)))
library(ggplot2)
g2 <- ggplot(data = stepsPerDay2, aes(stepsPerDay))
g2 <- g2 + geom_histogram()+labs(x="Total number of steps in a day", y = "Frequency", title = "Histogram 2 (missing values imputed)")
library(gridExtra)
grid.arrange(g1 + ylim(0, 12.5), g2 + ylim(0, 12.5), nrow = 2)
library(xtable)
summSteps2 <- summary(stepsPerDay2$stepsPerDay)
xt <- xtable(t(summSteps2))
caption(xt) = "Statistics for Daily Total Number of Steps (missing values imputed)"
print(xt, type = "html",include.rownames = FALSE)

```

From comparison with both histograms, it can be inferred that days with lots of missing data were having total steps close to zero, which it would mean that missing values were concentrated in specific days.  After missing values were imputed, there's more days with values greater than 10,000 steps than before.

The median for the steps variable in the new dataset  is `r formatC(summSteps2[3], format = "d", big.mark = ",")` daily total steps and the mean is `r formatC(summSteps2[4],format = "d", big.mark = ",")` daily total steps, which are both greater than the corresponding values for the original data with missing values not imputed.


## Are there differences in activity patterns between weekdays and weekends?


```{r weekd}

Sys.setlocale("LC_TIME", "English_United States.1252")
weekd <- c("Monday","Tuesday","Wednesday","Thursday","Friday")
activityData2$wkd = ifelse(is.element(weekdays(activityData2$date),weekd),"weekday","weekend")
table(activityData2$wkd)
meanStepsWeekday = data.frame(mean = with(activityData2[activityData2$wkd == "weekday",], 
                                           tapply(steps, interval, mean)))
meanStepsWeekday$interval = as.integer(rownames(meanStepsWeekday$mean))
meanStepsWeekday$wkd = factor(rep("weekday"))
meanStepsWeekend = data.frame(mean = with(activityData2[activityData2$wkd == "weekend",], 
                                           tapply(steps, interval, mean)))
meanStepsWeekend$interval = as.integer(rownames(meanStepsWeekend$mean))
meanStepsWeekend$wkd = factor(rep("weekend"))
meanSteps= rbind(meanStepsWeekday,meanStepsWeekend)
ggplot(data = meanSteps) + geom_line(aes(x = interval , y = mean, group = wkd, 
    colour = wkd ), size = 1) + facet_wrap(~wkd, ncol = 1, scales = "free_y") + 
    theme_bw() + theme(legend.position = "none") + labs(y = "Number of steps")
```

A new factor variable was created to indicate whether data was recorded on a weekday or a weekend day.  This variable was used to create daily activity patterns for intervals.  The average number of steps was calculated for each interval for weekdays and for weekend days and presented in the graphs below.  From this graphs, it might be inferred that during weekends, activity starts later than in weekdays, having peaks at around the same time, but the average activity is greater for weekend days and lasts a little later than in weekdays.  

